---
title: "Pollution and Gender Book Chapter: Analyis"
author: "Charlotte Sophia Bez and Linnea Nelli"
date: "`r Sys.Date()`"
output: html_document
description: produces all figures and maps shown in the book chapter.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 14, fig.height = 8)

setwd("C:\\Users\\charlott\\Dropbox (Personal)\\Book_Chapters\\pollution_gender")

library(readxl)
library(dplyr)
library(here)
library(tidyverse)
library(janitor)
library(sp)
library(sf)
library(readr)
library(biscale)
library(ggplot2)
library(cowplot)
library(purrr)
library(writexl)
library(lemon)
library(tidyr)
```

---

## 1. Inactivity Maps

Counties where **female inactivity** and **RSEI toxicity** are both in the top quintile (20%) or top decile (10%) are classified as bi-variate hotspots (`bi_class = "2-2"`). The loop produces individual threshold maps, which are then combined into a single overlay.

```{r inactivity-data}
empl_stat  <- read.csv("in/empl_stat.csv")
county_shp <- read_sf("in/tl_2022_us_county/tl_2022_us_county.shp")

TERRITORIES <- c("60", "66", "69", "72", "78")

county_geom <- county_shp %>%
  distinct(GEOID, STATEFP, geometry) %>%
  filter(!(STATEFP %in% TERRITORIES)) %>%
  arrange(GEOID)

state_shp <- read_sf("in/tl_2022_us_state/tl_2022_us_state.shp")
states <- state_shp %>%
  filter(!(STATEFP %in% c("60", "66", "69", "78", "72", "02", "15")))
```

```{r inactivity-loop}
for (thresh in list(list(q = 0.80, label = "20pc", pct = "20%"),
                    list(q = 0.90, label = "10pc", pct = "10%"))) {

  p_inactive <- quantile(empl_stat$share_Inactive_F, thresh$q, na.rm = TRUE)
  p_tox      <- quantile(empl_stat$Score_total,       thresh$q, na.rm = TRUE)

  inactive <- empl_stat %>%
    mutate(
      bi_class = case_when(
        share_Inactive_F >= p_inactive & Score_total >= p_tox ~ "2-2",
        share_Inactive_F >= p_inactive & Score_total <  p_tox ~ "2-1",
        share_Inactive_F <  p_inactive & Score_total >= p_tox ~ "1-2",
        TRUE ~ "1-1"
      )
    ) %>%
    group_by(STATE_COUNTY_FIP) %>%
    mutate(
      bi_class = case_when(
        any(bi_class == "2-2") ~ "2-2",
        any(bi_class == "2-1") ~ "2-1",
        any(bi_class == "1-2") ~ "1-2",
        TRUE ~ NA_character_
      )
    ) %>%
    ungroup() %>%
    arrange(STATE_COUNTY_FIP) %>%
    mutate(STATE_COUNTY_FIP = as.character(STATE_COUNTY_FIP)) %>%
    left_join(county_geom, by = c("STATE_COUNTY_FIP" = "GEOID"))

  inactive_states <- inactive %>%
    distinct(STATE_COUNTY_FIP, STATEFP, geometry, bi_class) %>%
    filter(!(STATEFP %in% c("02", "15", "72")), bi_class == "2-2")

  geom_states <- county_geom %>%
    filter(!(STATEFP %in% c("02", "15", "72")))

  inactive_states <- st_sf(inactive_states)
  geom_states     <- st_sf(geom_states)

  assign(paste0("inactive_states_", thresh$label), inactive_states)
  assign(paste0("geom_states_", thresh$label), geom_states)

  inactive_map <- ggplot() +
    geom_sf(data = states,          fill = "white", color = NA) +
    geom_sf(data = inactive_states, aes(fill = bi_class), color = NA,
            show.legend = FALSE) +
    geom_sf(data = states,          fill = NA, color = "grey50", linewidth = 0.3) +
    bi_scale_fill(pal = "DkViolet2", dim = 2, na.value = "white") +
    bi_theme() +
    coord_sf(expand = FALSE)

  ggsave(paste0("out/maps/inactive_tract_county_", thresh$label, ".png"),
         plot = inactive_map, dpi = 300)
  print(inactive_map)
}
```

```{r inactivity-combined-map}
combined_map <- ggplot() +
  geom_sf(data = states,               fill = "white", color = NA) +
  geom_sf(data = inactive_states_20pc, aes(fill = "20% hotspot counties in terms of inactivity"), color = NA) +
  geom_sf(data = inactive_states_10pc, aes(fill = "10% hotspot counties in terms of inactivity"), color = NA) +
  geom_sf(data = states,               fill = NA, color = "grey50", linewidth = 0.3) +
  scale_fill_manual(
    name   = NULL,
    values = c("20% hotspot counties in terms of inactivity" = "#1A0033",
               "10% hotspot counties in terms of inactivity" = "#7B0000"),
    guide  = guide_legend(override.aes = list(color = NA))
  ) +
  bi_theme() +
  coord_sf(expand = FALSE) +
  theme(
    legend.position      = c(0.03, 0.05),
    legend.justification = c(0, 0),
    legend.background    = element_rect(fill = "white", color = "grey60", linewidth = 0.4),
    legend.text          = element_text(size = 8),
    legend.key.size      = unit(0.32, "cm")
  )

ggsave("out/maps/inactive_tract_county_10pc_20pc.png", plot = combined_map, dpi = 300)
print(combined_map)
```

```{r inactivity-export}
counties_10pc <- inactive %>%
  group_by(STATE_COUNTY_FIP, CountyName, State_fips) %>%
  summarise(
    share_Inactive_F_mean = mean(share_Inactive_F, na.rm = TRUE),
    Score_total_mean      = mean(Score_total,       na.rm = TRUE),
    bi_class              = first(bi_class),
    .groups = "drop"
  ) %>%
  filter(bi_class == "2-2") %>%
  arrange(State_fips, CountyName)

writexl::write_xlsx(counties_10pc, "out/validation/counties_10pc_hotspots.xlsx")
cat("Exported", nrow(counties_10pc), "counties to out/counties_10pc_hotspots.xlsx\n")
```

```{r inactivity-county-list}
inactivity_hotspot_counties <- bind_rows(
    inactive_states_20pc %>% st_drop_geometry() %>% distinct(STATE_COUNTY_FIP),
    inactive_states_10pc %>% st_drop_geometry() %>% distinct(STATE_COUNTY_FIP)
  ) %>%
  distinct(STATE_COUNTY_FIP) %>%
  left_join(
    inactive %>% distinct(STATE_COUNTY_FIP, CountyName, State_fips),
    by = "STATE_COUNTY_FIP"
  ) %>%
  arrange(State_fips, CountyName)

cat("\n=== Counties in inactivity_tract_county_10pc_20pc map ===\n")
cat(paste0("  ", inactivity_hotspot_counties$CountyName,
           " (FIPS ", inactivity_hotspot_counties$STATE_COUNTY_FIP, ")"),
    sep = "\n")
cat(sprintf("\nTotal: %d counties\n", nrow(inactivity_hotspot_counties)))
```

---

## 2. Median Earnings Maps

Same bi-variate approach using inverted female earnings (`Female_inv`) instead of inactivity.

```{r earnings-data}
earnings <- read.csv("in/mean_hours.csv")

county_shp <- read_sf("in/tl_2022_us_county/tl_2022_us_county.shp")

county_geom <- county_shp %>%
  distinct(GEOID, STATEFP, geometry) %>%
  filter(STATEFP %in% unique(states_poll)) %>%
  arrange(GEOID)
```

```{r earnings-loop}
for (thresh in list(list(q = 0.80, label = "20pc", pct = "20%"),
                    list(q = 0.90, label = "10pc", pct = "10%"))) {

  p_earn_inv <- quantile(earnings$Female_inv, thresh$q, na.rm = TRUE)
  p_tox      <- quantile(earnings$Score_total, thresh$q, na.rm = TRUE)

  earnings_cl <- earnings %>%
    mutate(
      bi_class = case_when(
        Female_inv >= p_earn_inv & Score_total >= p_tox ~ "2-2",
        Female_inv >= p_earn_inv & Score_total <  p_tox ~ "2-1",
        Female_inv <  p_earn_inv & Score_total >= p_tox ~ "1-2",
        TRUE ~ "1-1"
      )
    ) %>%
    group_by(STATE_COUNTY_FIP) %>%
    mutate(
      bi_class = case_when(
        any(bi_class == "2-2") ~ "2-2",
        any(bi_class == "2-1") ~ "2-1",
        any(bi_class == "1-2") ~ "1-2",
        TRUE ~ NA_character_
      )
    ) %>%
    ungroup() %>%
    arrange(STATE_COUNTY_FIP) %>%
    mutate(STATE_COUNTY_FIP = as.character(STATE_COUNTY_FIP)) %>%
    left_join(county_geom, by = c("STATE_COUNTY_FIP" = "GEOID"))

  earnings_states <- earnings_cl %>%
    distinct(STATE_COUNTY_FIP, STATEFP, geometry, bi_class) %>%
    filter(!(STATEFP %in% c("02", "15", "72")), bi_class == "2-2")

  geom_earn_states <- county_geom %>%
    filter(!(STATEFP %in% c("02", "15", "72")))

  earnings_states  <- st_sf(earnings_states)
  geom_earn_states <- st_sf(geom_earn_states)

  assign(paste0("earnings_states_", thresh$label), earnings_states)

  earnings_map <- ggplot() +
    geom_sf(data = states,          fill = "white", color = NA) +
    geom_sf(data = earnings_states, aes(fill = bi_class), color = NA,
            show.legend = FALSE) +
    geom_sf(data = states,          fill = NA, color = "grey50", linewidth = 0.3) +
    bi_scale_fill(pal = "DkViolet2", dim = 2, na.value = "white") +
    bi_theme() +
    coord_sf(expand = FALSE)

  ggsave(paste0("out/maps/earnings_tract_county_", thresh$label, ".png"),
         plot = earnings_map, dpi = 300)
  print(earnings_map)
}
```

```{r earnings-combined-map}
earnings_combined_map <- ggplot() +
  geom_sf(data = states,               fill = "white", color = NA) +
  geom_sf(data = earnings_states_20pc, aes(fill = "20% hotspot counties in terms of earnings"), color = NA) +
  geom_sf(data = earnings_states_10pc, aes(fill = "10% hotspot counties in terms of earnings"), color = NA) +
  geom_sf(data = states,               fill = NA, color = "grey50", linewidth = 0.3) +
  scale_fill_manual(
    name   = NULL,
    values = c("20% hotspot counties in terms of earnings" = "#4169E1",
               "10% hotspot counties in terms of earnings" = "#00005C"),
    guide  = guide_legend(override.aes = list(color = NA))
  ) +
  bi_theme() +
  coord_sf(expand = FALSE) +
  theme(
    legend.position      = c(0.03, 0.05),
    legend.justification = c(0, 0),
    legend.background    = element_rect(fill = "white", color = "grey60", linewidth = 0.4),
    legend.text          = element_text(size = 8),
    legend.key.size      = unit(0.32, "cm")
  )

ggsave("out/maps/earnings_tract_county_10pc_20pc.png", plot = earnings_combined_map, dpi = 300)
print(earnings_combined_map)
```

```{r earnings-county-list}
earnings_hotspot_counties <- bind_rows(
    earnings_states_20pc %>% st_drop_geometry() %>% distinct(STATE_COUNTY_FIP),
    earnings_states_10pc %>% st_drop_geometry() %>% distinct(STATE_COUNTY_FIP)
  ) %>%
  distinct(STATE_COUNTY_FIP) %>%
  left_join(
    earnings_cl %>% distinct(STATE_COUNTY_FIP, CountyName, State_fips),
    by = "STATE_COUNTY_FIP"
  ) %>%
  arrange(State_fips, CountyName)

cat("\n=== Counties in earnings_tract_county_10pc_20pc map ===\n")
cat(paste0("  ", earnings_hotspot_counties$CountyName,
           " (FIPS ", earnings_hotspot_counties$STATE_COUNTY_FIP, ")"),
    sep = "\n")
cat(sprintf("\nTotal: %d counties\n", nrow(earnings_hotspot_counties)))
```

---

## 3. Employment Maps

```{r employment-loop}
empl_stat <- read.csv("in/empl_stat.csv")

for (thresh in list(list(q = 0.80, label = "20pc", pct = "20%"),
                    list(q = 0.90, label = "10pc", pct = "10%"))) {

  p_employ <- quantile(empl_stat$share_Employed_F, thresh$q, na.rm = TRUE)
  p_tox    <- quantile(empl_stat$Score_total,       thresh$q, na.rm = TRUE)

  employ_cl <- empl_stat %>%
    mutate(
      bi_class = case_when(
        share_Employed_F >= p_employ & Score_total >= p_tox ~ "2-2",
        share_Employed_F >= p_employ & Score_total <  p_tox ~ "2-1",
        share_Employed_F <  p_employ & Score_total >= p_tox ~ "1-2",
        TRUE ~ "1-1"
      )
    ) %>%
    group_by(STATE_COUNTY_FIP) %>%
    mutate(
      bi_class = case_when(
        any(bi_class == "2-2") ~ "2-2",
        any(bi_class == "2-1") ~ "2-1",
        any(bi_class == "1-2") ~ "1-2",
        TRUE ~ NA_character_
      )
    ) %>%
    ungroup() %>%
    arrange(STATE_COUNTY_FIP) %>%
    mutate(STATE_COUNTY_FIP = as.character(STATE_COUNTY_FIP)) %>%
    left_join(county_geom, by = c("STATE_COUNTY_FIP" = "GEOID"))

  employ_states <- employ_cl %>%
    distinct(STATE_COUNTY_FIP, STATEFP, geometry, bi_class) %>%
    filter(!(STATEFP %in% c("02", "15", "72")), bi_class == "2-2")

  employ_states <- st_sf(employ_states)
  assign(paste0("employ_states_", thresh$label), employ_states)

  employ_map <- ggplot() +
    geom_sf(data = states,        fill = "white", color = NA) +
    geom_sf(data = employ_states, aes(fill = bi_class), color = NA,
            show.legend = FALSE) +
    geom_sf(data = states,        fill = NA, color = "grey50", linewidth = 0.3) +
    bi_scale_fill(pal = "DkViolet2", dim = 2, na.value = "white") +
    bi_theme() +
    coord_sf(expand = FALSE)

  ggsave(paste0("out/maps/employ_tract_county_", thresh$label, ".png"),
         plot = employ_map, dpi = 300)
  print(employ_map)
}
```

```{r employment-combined-map}
employ_combined_map <- ggplot() +
  geom_sf(data = states,             fill = "white", color = NA) +
  geom_sf(data = employ_states_20pc, aes(fill = "20% hotspot counties in terms of employment"), color = NA) +
  geom_sf(data = employ_states_10pc, aes(fill = "10% hotspot counties in terms of employment"), color = NA) +
  geom_sf(data = states,             fill = NA, color = "grey50", linewidth = 0.3) +
  scale_fill_manual(
    name   = NULL,
    values = c("20% hotspot counties in terms of employment" = "#D4820A",
               "10% hotspot counties in terms of employment" = "#6B3A00"),
    guide  = guide_legend(override.aes = list(color = NA))
  ) +
  bi_theme() +
  coord_sf(expand = FALSE) +
  theme(
    legend.position      = c(0.03, 0.05),
    legend.justification = c(0, 0),
    legend.background    = element_rect(fill = "white", color = "grey60", linewidth = 0.4),
    legend.text          = element_text(size = 5),
    legend.key.size      = unit(0.17, "cm")
  )

ggsave("out/maps/employ_tract_county_10pc_20pc.png", plot = employ_combined_map, dpi = 300)
print(employ_combined_map)
```

---

## 4. Toxicity-Only Map

```{r toxicity-map}
for (thresh in list(list(q = 0.80, label = "20pc"),
                    list(q = 0.90, label = "10pc"))) {

  p_tox <- quantile(empl_stat$Score_total, thresh$q, na.rm = TRUE)

  tox <- empl_stat %>%
    mutate(hotspot = Score_total >= p_tox) %>%
    group_by(STATE_COUNTY_FIP) %>%
    mutate(hotspot = any(hotspot)) %>%
    ungroup() %>%
    arrange(STATE_COUNTY_FIP) %>%
    mutate(STATE_COUNTY_FIP = as.character(STATE_COUNTY_FIP)) %>%
    left_join(county_geom, by = c("STATE_COUNTY_FIP" = "GEOID"))

  tox_states <- tox %>%
    distinct(STATE_COUNTY_FIP, STATEFP, geometry, hotspot) %>%
    filter(!(STATEFP %in% c("02", "15", "72")), hotspot == TRUE)

  assign(paste0("tox_states_", thresh$label), st_sf(tox_states))
}

tox_geom <- county_geom %>%
  filter(!(STATEFP %in% c("02", "15", "72"))) %>%
  st_sf()

tox_combined_map <- ggplot() +
  geom_sf(data = states,          fill = "white", color = NA) +
  geom_sf(data = tox_states_20pc, aes(fill = "20% hotspot counties"), color = NA) +
  geom_sf(data = tox_states_10pc, aes(fill = "10% hotspot counties"), color = NA) +
  geom_sf(data = states,          fill = NA, color = "grey50", linewidth = 0.3) +
  scale_fill_manual(
    name   = NULL,
    values = c("20% hotspot counties" = "#2E8B57", "10% hotspot counties" = "#004D00"),
    guide  = guide_legend(override.aes = list(color = NA))
  ) +
  coord_sf(expand = FALSE) +
  theme_void() +
  theme(
    legend.position      = c(0.03, 0.05),
    legend.justification = c(0, 0),
    legend.background    = element_rect(fill = "white", color = "grey60", linewidth = 0.4),
    legend.text          = element_text(size = 5),
    legend.key.size      = unit(0.17, "cm")
  )

ggsave("out/maps/toxicity_tract_county_10pc_20pc.png", plot = tox_combined_map, dpi = 300)
print(tox_combined_map)
```

---

## 5. Solar Facilities Map

```{r solar-data}
solar_data_map <- read.csv("in/solar_data_map.csv", sep = ";")

clean_coordinate <- function(x, min_val, max_val) {
  x_num <- as.numeric(gsub("[^0-9-]", "", x))
  if (is.na(x_num)) return(NA)
  val          <- abs(x_num)
  lower_bound  <- min(abs(min_val), abs(max_val))
  upper_bound  <- max(abs(min_val), abs(max_val))
  while (val > upper_bound) val <- val / 10
  return(ifelse(x_num < 0, -val, val))
}

solar_geo <- solar_data_map %>%
  rowwise() %>%
  mutate(
    lon_clean = clean_coordinate(longitude, -125, -66),
    lat_clean = clean_coordinate(latitude,   24,   49)
  ) %>%
  ungroup() %>%
  filter(!is.na(lon_clean), !is.na(lat_clean))

solar_sf <- st_as_sf(solar_geo, coords = c("lon_clean", "lat_clean"), crs = 4326)
```

```{r solar-map}
ggplot() +
  geom_sf(data = states,   fill = "white", color = "grey80", size = 0.1) +
  geom_sf(data = solar_sf, aes(color = facility_type), size = 2, alpha = 0.7) +
  scale_color_viridis_d(option = "plasma") +
  coord_sf(expand = FALSE) +
  theme_void() +
  labs(color = "Facility Type")

ggsave("out/maps/solar_map.png", dpi = 300)
```

### 5.1 Solar overlay – national (inactivity hotspots)

```{r solar-hotspot-buffer}
hotspot_buffer <- inactive_states_20pc %>%
  st_transform(5070) %>%
  st_union() %>%
  st_buffer(16093)

solar_near_hotspots <- solar_sf %>%
  st_transform(5070) %>%
  st_filter(hotspot_buffer) %>%
  st_transform(st_crs(geom_states_10pc))
```

```{r solar-overlay-national}
overlay_map <- ggplot() +
  geom_sf(data = states,               fill = "white", color = NA) +
  geom_sf(data = inactive_states_20pc, aes(fill = "20% hotspot counties"), color = NA) +
  geom_sf(data = inactive_states_10pc, aes(fill = "10% hotspot counties"), color = NA) +
  scale_fill_manual(
    name   = NULL,
    values = c("20% hotspot counties" = "#1A0033", "10% hotspot counties" = "#7B0000"),
    guide  = guide_legend(override.aes = list(color = NA))
  ) +
  geom_sf(data = solar_near_hotspots, aes(color = facility_type), size = 1.5, alpha = 0.85) +
  geom_sf(data = states, fill = NA, color = "grey50", linewidth = 0.3) +
  scale_color_viridis_d(option = "plasma", name = "Solar facility type") +
  coord_sf(expand = FALSE) +
  theme_void() +
  theme(
    legend.position = "bottom",
    legend.title    = element_text(face = "bold", size = 5),
    legend.text     = element_text(size = 5)
  ) +
  guides(
    fill  = guide_legend(order = 1, override.aes = list(color = NA)),
    color = guide_legend(order = 2, nrow = 2, override.aes = list(size = 3))
  )

ggsave("out/maps/solar_inactive_overlay.png", plot = overlay_map, width = 14, height = 8, dpi = 300)
print(overlay_map)
```

### 5.2 Solar overlay – Rust Belt

```{r solar-overlay-rustbelt}
xmin <- -92.0; xmax <- -74.5; ymin <- 36.5; ymax <- 47.5

overlay_map_Rust_Belt <- ggplot() +
  geom_sf(data = states,               fill = "white", color = NA) +
  geom_sf(data = inactive_states_20pc, aes(fill = "20% hotspot counties"), color = NA) +
  geom_sf(data = inactive_states_10pc, aes(fill = "10% hotspot counties"), color = NA) +
  scale_fill_manual(
    name   = NULL,
    values = c("20% hotspot counties" = "#1A0033", "10% hotspot counties" = "#7B0000"),
    guide  = guide_legend(override.aes = list(color = NA))
  ) +
  geom_sf(data = solar_near_hotspots, aes(color = facility_type), size = 1.5, alpha = 0.85) +
  geom_sf(data = states, fill = NA, color = "grey50", linewidth = 0.3) +
  scale_color_viridis_d(option = "plasma", name = "Solar facility type") +
  coord_sf(xlim = c(xmin, xmax), ylim = c(ymin, ymax), expand = FALSE) +
  theme_void() +
  theme(legend.position = "bottom",
        legend.title    = element_text(face = "bold", size = 5),
        legend.text     = element_text(size = 5)) +
  guides(fill  = guide_legend(order = 1, override.aes = list(color = NA)),
         color = guide_legend(order = 2, nrow = 2, override.aes = list(size = 3)))

ggsave("out/maps/rustbelt/solar_inactive_overlay_Rust_Belt.png",
       plot = overlay_map_Rust_Belt, width = 10, height = 7, dpi = 300)
print(overlay_map_Rust_Belt)
```

### 5.3 Solar overlay – South

```{r solar-overlay-south}
xmin <- -97.5; xmax <- -85.5; ymin <- 28.5; ymax <- 35.5

overlay_map_South <- ggplot() +
  geom_sf(data = states,               fill = "white", color = NA) +
  geom_sf(data = inactive_states_20pc, aes(fill = "20% hotspot counties"), color = NA) +
  geom_sf(data = inactive_states_10pc, aes(fill = "10% hotspot counties"), color = NA) +
  scale_fill_manual(
    name   = NULL,
    values = c("20% hotspot counties" = "#1A0033", "10% hotspot counties" = "#7B0000"),
    guide  = guide_legend(override.aes = list(color = NA))
  ) +
  geom_sf(data = solar_near_hotspots, aes(color = facility_type), size = 1.5, alpha = 0.85) +
  geom_sf(data = states, fill = NA, color = "grey50", linewidth = 0.3) +
  scale_color_viridis_d(option = "plasma", name = "Solar facility type") +
  coord_sf(xlim = c(xmin, xmax), ylim = c(ymin, ymax), expand = FALSE) +
  theme_void() +
  theme(legend.position = "bottom",
        legend.title    = element_text(face = "bold", size = 5),
        legend.text     = element_text(size = 5)) +
  guides(fill  = guide_legend(order = 1, override.aes = list(color = NA)),
         color = guide_legend(order = 2, nrow = 2, override.aes = list(size = 3)))

ggsave("out/maps/south/solar_inactive_overlay_South.png",
       plot = overlay_map_South, width = 10, height = 7, dpi = 300)
print(overlay_map_South)
```

### 5.4 Solar overlay – Utah (no legend)

```{r solar-overlay-utah}
xmin <- -114.1; xmax <- -109.0; ymin <- 36.9; ymax <- 42.0

overlay_map_Utah <- ggplot() +
  geom_sf(data = states,               fill = "white", color = NA) +
  geom_sf(data = inactive_states_20pc, aes(fill = "20% hotspot counties"), color = NA) +
  geom_sf(data = inactive_states_10pc, aes(fill = "10% hotspot counties"), color = NA) +
  scale_fill_manual(
    name   = NULL,
    values = c("20% hotspot counties" = "#1A0033", "10% hotspot counties" = "#7B0000"),
    guide  = "none"
  ) +
  geom_sf(data = solar_near_hotspots, aes(color = facility_type), size = 1.5, alpha = 0.85) +
  scale_color_viridis_d(option = "plasma", name = "Solar facility type", guide = "none") +
  coord_sf(xlim = c(xmin, xmax), ylim = c(ymin, ymax), expand = FALSE) +
  theme_void()

ggsave("out/maps/utah/solar_inactive_overlay_Utah.png",
       plot = overlay_map_Utah, width = 10, height = 7, dpi = 300)
print(overlay_map_Utah)
```

---

## 6. Battery Supply-Chain Facilities Map

```{r battery-data}
file_path      <- "in/naatbatt-database-31march2026.xlsx"
all_sheets     <- excel_sheets(file_path)
target_sheets  <- all_sheets[8:19]

battery_facilities <- target_sheets %>%
  set_names() %>%
  map_df(~read_excel(path = file_path, sheet = .x,
                     col_types = "text", trim_ws = TRUE),
         .id = "origin_sheet")

battery_geo <- battery_facilities %>%
  mutate(
    lon_clean        = as.numeric(Longitude),
    lat_clean        = as.numeric(Latitude),
    missing          = ifelse(is.na(lon_clean) & !is.na(Longitude), 1, 0),
    facility_country = ifelse(`Facility State or Province` == "ON", "Canada", `Facility Country`),
    lon_clean        = ifelse(lon_clean > 0, lon_clean * -1, lon_clean)
  ) %>%
  filter(is.finite(lon_clean) & is.finite(lat_clean),
         `Facility Country` == "US",
         !(`Facility State or Province` %in% c("TBA", "AK"))) %>%
  mutate(
    lon_clean = ifelse(lon_clean == -70.60000 & lat_clean == 40.00000, -74.7126, lon_clean),
    lat_clean = ifelse(lon_clean == -70.60000 & lat_clean == 40.00000,  40.1135, lat_clean)
  )

battery_sf <- st_as_sf(battery_geo, coords = c("lon_clean", "lat_clean"), crs = 4326)
battery_sf$origin_sheet <- factor(battery_sf$origin_sheet,
                                   levels = unique(battery_sf$origin_sheet))
```

```{r battery-overlay-national}
battery_near_hotspots <- battery_sf %>%
  st_transform(5070) %>%
  st_filter(hotspot_buffer) %>%
  st_transform(st_crs(geom_states_10pc))

battery_overlay_map <- ggplot() +
  geom_sf(data = states,               fill = "white", color = NA) +
  geom_sf(data = inactive_states_20pc, aes(fill = "20% hotspot counties"), color = NA) +
  geom_sf(data = inactive_states_10pc, aes(fill = "10% hotspot counties"), color = NA) +
  scale_fill_manual(
    name   = NULL,
    values = c("20% hotspot counties" = "#1A0033", "10% hotspot counties" = "#7B0000"),
    guide  = guide_legend(override.aes = list(color = NA))
  ) +
  geom_sf(data = battery_near_hotspots, aes(color = origin_sheet), size = 1.5, alpha = 0.85) +
  geom_sf(data = states, fill = NA, color = "grey50", linewidth = 0.3) +
  scale_color_viridis_d(option = "plasma", name = "Battery facility type") +
  coord_sf(expand = FALSE) +
  theme_void() +
  theme(legend.position = "bottom",
        legend.title    = element_text(face = "bold", size = 8),
        legend.text     = element_text(size = 8)) +
  guides(fill  = guide_legend(order = 1, override.aes = list(color = NA)),
         color = guide_legend(order = 2, nrow = 2, override.aes = list(size = 3)))

ggsave("out/maps/battery_inactive_overlay.png",
       plot = battery_overlay_map, width = 14, height = 8, dpi = 300)
print(battery_overlay_map)
```

### 6.1 Battery overlay – Rust Belt

```{r battery-overlay-rustbelt}
xmin <- -92.0; xmax <- -74.5; ymin <- 36.5; ymax <- 47.5

battery_overlay_map_Rust_Belt <- ggplot() +
  geom_sf(data = states,               fill = "white", color = NA) +
  geom_sf(data = inactive_states_20pc, aes(fill = "20% hotspot counties"), color = NA) +
  geom_sf(data = inactive_states_10pc, aes(fill = "10% hotspot counties"), color = NA) +
  scale_fill_manual(
    name   = NULL,
    values = c("20% hotspot counties" = "#1A0033", "10% hotspot counties" = "#7B0000"),
    guide  = guide_legend(override.aes = list(color = NA))
  ) +
  geom_sf(data = battery_near_hotspots, aes(color = origin_sheet), size = 1.5, alpha = 0.85) +
  geom_sf(data = states, fill = NA, color = "grey50", linewidth = 0.3) +
  scale_color_viridis_d(option = "plasma", name = "Battery facility type") +
  coord_sf(xlim = c(xmin, xmax), ylim = c(ymin, ymax), expand = FALSE) +
  theme_void() +
  theme(legend.position = "bottom",
        legend.title    = element_text(face = "bold", size = 8),
        legend.text     = element_text(size = 7)) +
  guides(fill  = guide_legend(order = 1, override.aes = list(color = NA)),
         color = guide_legend(order = 2, nrow = 2, override.aes = list(size = 3)))

ggsave("out/maps/rustbelt/battery_inactive_overlay_rustbelt.png",
       plot = battery_overlay_map_Rust_Belt, width = 10, height = 7, dpi = 300)
print(battery_overlay_map_Rust_Belt)
```

### 6.2 Battery overlay – South

```{r battery-overlay-south}
xmin <- -97.5; xmax <- -85.5; ymin <- 28.5; ymax <- 35.5

battery_overlay_map_South <- ggplot() +
  geom_sf(data = states,               fill = "white", color = NA) +
  geom_sf(data = inactive_states_20pc, aes(fill = "20% hotspot counties"), color = NA) +
  geom_sf(data = inactive_states_10pc, aes(fill = "10% hotspot counties"), color = NA) +
  scale_fill_manual(
    name   = NULL,
    values = c("20% hotspot counties" = "#1A0033", "10% hotspot counties" = "#7B0000"),
    guide  = guide_legend(override.aes = list(color = NA))
  ) +
  geom_sf(data = battery_near_hotspots, aes(color = origin_sheet), size = 1.5, alpha = 0.85) +
  geom_sf(data = states, fill = NA, color = "grey50", linewidth = 0.3) +
  scale_color_viridis_d(option = "plasma", name = "Battery facility type") +
  coord_sf(xlim = c(xmin, xmax), ylim = c(ymin, ymax), expand = FALSE) +
  theme_void() +
  theme(legend.position = "bottom",
        legend.title    = element_text(face = "bold", size = 8),
        legend.text     = element_text(size = 7)) +
  guides(fill  = guide_legend(order = 1, override.aes = list(color = NA)),
         color = guide_legend(order = 2, nrow = 2, override.aes = list(size = 3)))

ggsave("out/maps/south/battery_inactive_overlay_South.png",
       plot = battery_overlay_map_South, width = 10, height = 7, dpi = 300)
print(battery_overlay_map_South)
```

### 6.3 Battery overlay – Utah (no legend)

```{r battery-overlay-utah}
xmin <- -114.1; xmax <- -109.0; ymin <- 36.9; ymax <- 42.0

battery_overlay_map_Utah <- ggplot() +
  geom_sf(data = states,               fill = "white", color = NA) +
  geom_sf(data = inactive_states_20pc, aes(fill = "20% hotspot counties"), color = NA) +
  geom_sf(data = inactive_states_10pc, aes(fill = "10% hotspot counties"), color = NA) +
  scale_fill_manual(
    name   = NULL,
    values = c("20% hotspot counties" = "#1A0033", "10% hotspot counties" = "#7B0000"),
    guide  = "none"
  ) +
  geom_sf(data = battery_near_hotspots, aes(color = origin_sheet), size = 1.5, alpha = 0.85) +
  scale_color_viridis_d(option = "plasma", name = "Battery facility type", guide = "none") +
  coord_sf(xlim = c(xmin, xmax), ylim = c(ymin, ymax), expand = FALSE) +
  theme_void()

ggsave("out/maps/utah/battery_inactive_overlay_Utah.png",
       plot = battery_overlay_map_Utah, width = 10, height = 7, dpi = 300)
print(battery_overlay_map_Utah)
```

### 6.4 Full battery map with Alaska inset

```{r battery-full-map}
states_ak   <- state_shp %>% filter(STATEFP == "02")

battery_ak <- battery_facilities %>%
  mutate(
    lon_clean        = as.numeric(Longitude),
    lat_clean        = as.numeric(Latitude),
    missing          = ifelse(is.na(lon_clean) & !is.na(Longitude), 1, 0),
    facility_country = ifelse(`Facility State or Province` == "ON", "Canada", `Facility Country`),
    lon_clean        = ifelse(lon_clean > 0, lon_clean * -1, lon_clean)
  ) %>%
  filter(is.finite(lon_clean) & is.finite(lat_clean),
         `Facility Country` == "US",
         `Facility State or Province` == "AK")

battery_ak_sf <- st_as_sf(battery_ak, coords = c("lon_clean", "lat_clean"), crs = 4326)

ak_map <- ggplot() +
  geom_sf(data = states_ak,      fill = "white", color = "grey80", size = 0.1) +
  geom_sf(data = battery_ak_sf,  aes(color = origin_sheet),
          size = 2, alpha = 0.7, show.legend = FALSE) +
  coord_sf(xlim = c(-170, -130), ylim = c(50, 72), expand = FALSE) +
  theme_void() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 0.5))

tot_map <- ggplot() +
  geom_sf(data = states,      fill = "white", color = "grey80", size = 0.1) +
  geom_sf(data = battery_sf,  aes(color = origin_sheet), size = 2, alpha = 0.7) +
  scale_color_discrete(drop = FALSE) +
  theme_void() +
  labs(color = "Facility Type") +
  guides(color = guide_legend(position = "right", nrow = 12))

battery_plot <- ggdraw() +
  draw_plot(tot_map, 0,    0,    1,    1) +
  draw_plot(ak_map,  0.05, 0.05, 0.20, 0.30)

ggsave("out/maps/battery_map.png", plot = battery_plot, dpi = 300)
print(battery_plot)
```

---

## 7. Co-location Statistics

Share of inactivity/earnings hotspot counties that have at least one solar or battery facility within **10 miles**.

```{r colocation-stats}
solar_proj   <- solar_sf   %>% st_transform(5070)
battery_proj <- battery_sf %>% st_transform(5070)

coloc_pct <- function(hotspot_sf, facility_proj) {
  buf <- hotspot_sf %>% st_transform(5070) %>% st_buffer(16093)
  mean(lengths(st_intersects(buf, facility_proj)) > 0, na.rm = TRUE) * 100
}

inact_n       <- nrow(inactive_states_20pc)
earn_n        <- nrow(earnings_states_20pc)

inact_solar   <- coloc_pct(inactive_states_20pc, solar_proj)
inact_battery <- coloc_pct(inactive_states_20pc, battery_proj)
earn_solar    <- coloc_pct(earnings_states_20pc,  solar_proj)
earn_battery  <- coloc_pct(earnings_states_20pc,  battery_proj)

cat("\n===== CO-LOCATION STATISTICS (facility within 10 miles of hotspot county) =====\n")
cat(sprintf("\nAmong the %d inactivity hotspot counties, %.1f%% have a solar facility within 10 miles.\n",
            inact_n, inact_solar))
cat(sprintf("Among inactivity hotspot counties, %.1f%% have a battery facility within 10 miles.\n",
            inact_battery))
cat(sprintf("\nAmong the %d low-earnings hotspot counties, %.1f%% have a solar facility within 10 miles.\n",
            earn_n, earn_solar))
cat(sprintf("Among low-earnings hotspot counties, %.1f%% have a battery facility within 10 miles.\n",
            earn_battery))
```

```{r colocation-county-lists}
inact_proj  <- inactive_states_20pc %>% st_transform(5070)
inact_buf   <- inact_proj %>% st_buffer(16093)

has_solar   <- lengths(st_intersects(inact_buf, solar_proj))   > 0
has_battery <- lengths(st_intersects(inact_buf, battery_proj)) > 0

inact_county_names <- inactive_states_20pc %>%
  st_drop_geometry() %>%
  left_join(inactive %>% distinct(STATE_COUNTY_FIP, CountyName, State_fips),
            by = "STATE_COUNTY_FIP")

solar_counties   <- inact_county_names[has_solar,   ] %>% arrange(State_fips, CountyName)
battery_counties <- inact_county_names[has_battery, ] %>% arrange(State_fips, CountyName)

cat("\n=== Inactivity hotspot counties with a SOLAR facility within 10 miles ===\n")
cat(paste0("  ", solar_counties$CountyName,
           " (FIPS ", solar_counties$STATE_COUNTY_FIP, ")"), sep = "\n")
cat(sprintf("\nTotal: %d counties\n", nrow(solar_counties)))

cat("\n=== Inactivity hotspot counties with a BATTERY facility within 10 miles ===\n")
cat(paste0("  ", battery_counties$CountyName,
           " (FIPS ", battery_counties$STATE_COUNTY_FIP, ")"), sep = "\n")
cat(sprintf("\nTotal: %d counties\n", nrow(battery_counties)))
```

```{r colocation-overlap}
inact_fips   <- inactive_states_20pc$STATE_COUNTY_FIP
earn_fips    <- earnings_states_20pc$STATE_COUNTY_FIP
n_overlap    <- length(intersect(inact_fips, earn_fips))
pct_of_inact <- n_overlap / length(inact_fips) * 100
pct_of_earn  <- n_overlap / length(earn_fips)  * 100

cat(sprintf(
  "\n%d counties appear in both hotspot sets (20%% threshold): %.1f%% of inactivity hotspots, %.1f%% of earnings hotspots.\n",
  n_overlap, pct_of_inact, pct_of_earn))
```

---

## 8. County Lists Export

Single Excel file with one sheet per map, listing individual census tracts visible in each figure.

```{r tract-lists-export}
make_sheet <- function(hotspot_fips_vec, tract_data, outcome_var) {
  tract_data %>%
    mutate(STATE_COUNTY_FIP = as.character(STATE_COUNTY_FIP)) %>%
    filter(STATE_COUNTY_FIP %in% as.character(hotspot_fips_vec),
           bi_class == "2-2") %>%
    mutate(tract_rank = dense_rank(desc(Score_total))) %>%
    arrange(tract_rank, State_fips, CountyName) %>%
    select(tract_rank, GEOID, State_fips, CountyName, STATE_COUNTY_FIP,
           all_of(outcome_var), Score_total)
}

# Extract hotspot FIPS from two sf threshold layers
hotspot_fips <- function(sf_20pc, sf_10pc) {
  bind_rows(
    sf_20pc %>% st_drop_geometry() %>% distinct(STATE_COUNTY_FIP),
    sf_10pc %>% st_drop_geometry() %>% distinct(STATE_COUNTY_FIP)
  ) %>% distinct(STATE_COUNTY_FIP) %>% pull(STATE_COUNTY_FIP)
}

# Extract hotspot FIPS restricted to a bounding box
hotspot_fips_bbox <- function(sf_nearby, xmin, xmax, ymin, ymax) {
  bbox_poly <- st_as_sfc(st_bbox(c(xmin = xmin, xmax = xmax,
                                    ymin = ymin, ymax = ymax),
                                  crs = st_crs(4326))) %>%
    st_transform(st_crs(sf_nearby))
  sf_nearby %>%
    st_filter(bbox_poly) %>%
    st_drop_geometry() %>%
    distinct(STATE_COUNTY_FIP) %>%
    pull(STATE_COUNTY_FIP)
}

# sf subsets: inactivity hotspot counties that have a facility within 10 miles
inact_with_battery <- inactive_states_20pc[has_battery, ]
inact_with_solar   <- inactive_states_20pc[has_solar,   ]

sheet1_earnings   <- make_sheet(hotspot_fips(earnings_states_20pc, earnings_states_10pc), earnings_cl, "Female")
sheet2_inactivity <- make_sheet(hotspot_fips(inactive_states_20pc, inactive_states_10pc), inactive,    "share_Inactive_F")

sheet3_battery_rb <- make_sheet(hotspot_fips_bbox(inact_with_battery, -92.0, -74.5, 36.5, 47.5), inactive, "share_Inactive_F")
sheet4_solar_rb   <- make_sheet(hotspot_fips_bbox(inact_with_solar,   -92.0, -74.5, 36.5, 47.5), inactive, "share_Inactive_F")

sheet5_battery_s  <- make_sheet(hotspot_fips_bbox(inact_with_battery, -97.5, -85.5, 28.5, 35.5), inactive, "share_Inactive_F")
sheet6_solar_s    <- make_sheet(hotspot_fips_bbox(inact_with_solar,   -97.5, -85.5, 28.5, 35.5), inactive, "share_Inactive_F")

sheet7_utah       <- make_sheet(hotspot_fips_bbox(inact_with_battery, -114.1, -109.0, 36.9, 42.0), inactive, "share_Inactive_F")

writexl::write_xlsx(
  list(
    "Earnings hotspots"          = sheet1_earnings,
    "Inactivity hotspots"        = sheet2_inactivity,
    "Battery overlay Rust Belt"  = sheet3_battery_rb,
    "Solar overlay Rust Belt"    = sheet4_solar_rb,
    "Battery overlay South"      = sheet5_battery_s,
    "Solar overlay South"        = sheet6_solar_s,
    "Utah hotspots"              = sheet7_utah
  ),
  "out/validation/county_lists_by_map.xlsx"
)
cat("Tract lists exported to out/validation/county_lists_by_map.xlsx\n")
```
